---
title: "Prepare data"
author: "Julia Romanowska"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Prepare_data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

This vignette will show you how to prepare the DNA methylation dataset that
you have read into the memory as shown in `Read data` vignette. The preparation
is necessary to use this methylation data together with genetic data in
interaction analyses in [`Haplin` package](https://cran.r-project.org/package=Haplin).

```{r setup}
library(HaplinMethyl)

ex_path <- system.file("extdata", package = "HaplinMethyl")
ex_file <- "env_data_test.dat"
ex_out_file <- "dnam_ex"

dnam_ex <- envDataRead(
  file.in = ex_file,
  dir.in = ex_path,
  file.out = ex_out_file,
  sep = " ",
  overwrite = TRUE
)
```

# Subsetting

Let's check again what does `dnam_ex` object include:

```{r}
dnam_ex
summary(dnam_ex)
```

When you don't want to use the entire data, you can use `envDataSubset` function
to easily specify which filters to apply to the dataset.

You can subset using row names, row numbers, column names, or column numbers.
This is useful when you want to, e.g., extract only the measurements from
one subgroup of samples or focus on specific CpGs.

> Check the `Finding CpGs` vignette on how to use the functions that find CpGs
> within a defined region!

```{r}
dnam_ex_3cpgs <- envDataSubset(
  env.data = dnam_ex,
  col.names = c("cg5", "cg7", "cg10"),
  file.out = "dnam_ex_3cpgs"
)
```

This produces two new files: `dnam_ex_3cpgs_gen.ffData` and
`dnam_ex_3cpgs_gen.RData`, which again can be used to load the data faster
in the next R-session instance.

And the returned object is of the same class as the original one, but has only
the three chosen CpGs:

```{r}
dnam_ex_3cpgs
summary(dnam_ex_3cpgs)
```

# Categorizing

When the measurements are continuous (as it is usually with data from arrays),
we need to first categorize it to use in `Haplin`. Thus, the level of DNA
methylation at a single CpG or a CpG region will dictate the stratum membership
for each sample.

```{r}
dnam_ex_3cpgs_cat <- envDataCategorize(
  env.data = dnam_ex_3cpgs,
  breaks = 3,
  file.out = "dnam_ex_3cpg_cat",
  overwrite = TRUE
)
dnam_ex_3cpgs_cat
class(dnam_ex_3cpgs_cat)
showRaw(dnam_ex_3cpgs_cat)
```

