---
title: "Reading DNA methylation data"
author: "Julia Romanowska"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Read_data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

This vignette gives an overview of `HaplinMethyl` functions to read DNA 
methylation data to memory.

# Read data from a file

Let's use the exemplary data file:

```{r setup}
library(HaplinMethyl)

ex_path <- system.file("extdata", package = "HaplinMethyl")
ex_file <- "env_data_test.dat"
ex_out_file <- "dnam_ex"
```

To read directly from a file, use `envDataRead` function:

```{r}
dnam_ex <- envDataRead(
  file.in = ex_file,
  dir.in = ex_path,
  file.out = ex_out_file,
  sep = " ", # the exemplary file is a space-delimited file
  header = TRUE, # make sure to check this!
  rownames = TRUE, # make sure to check this!
  overwrite = TRUE
)
```

Now, in your directory, you have new files:

```{r}
dir(pattern = ex_out_file)
```

> If you want to use this data in future analyses, don't delete those files!

# Read data from a matrix

It is also possible to create this object from a matrix that is already
in the memory.

```{r, eval=FALSE}
dnam_ex_from_matrix <- envDataReadFromObj(dnam_matrix)
```

# Loading data

Reading in our exemplary dataset did not take long time as this data is small.
However, with real-world data, this might take even up to an hour. Thus, you
don't want to do it every time you start up a new analysis!

If you close your current R-session or delete the DNA methylation object, you
can easily re-load it by using `envDataLoad` function and the new files 
created by `envDataRead`.

```{r, eval=FALSE}
dnam_ex <- envDataLoad(ex_out_file)
```

> **NOTE:** loading the data from .ffData file takes seconds instead of tens 
> of minutes! :)

All of the mentioned functions give an object of class `env.data` and either
`env.cont` for continuous measurements or `env.cat` for categorical data.

```{r}
class(dnam_ex)
```


# Look at data

DNA methylation data is usually stored as a huge matrix with CpGs in columns
and samples in rows, or vice-versa. Each cell of the matrix stores a $\beta$
value, which tells us the rate of methylation of the CpG in the sample (a
number between 0 and 1). It might be difficult to print the entire dataset,
so we've created some help functions.

```{r}
dnam_ex
summary(dnam_ex)
```

If you want the complete vector of row names and/or column names, use
`short = FALSE` argument of the `summary` function:

```{r}
summary(dnam_ex, short = FALSE)
```

You can also get quickly number of rows and columns:

```{r}
nrows(dnam_ex)
ncolumns(dnam_ex)
```


